---
title: "Data Prep: Predicting Type 2 Diabetes Onset from Longitudinal EHR Data in an Asian American Cohort"
author: "Veda Sripada"
date: "`r Sys.Date()`"
output: pdf_document
params:
  data_dir: "data/raw"        # folder holding the .rdata extracts (not in this repository)
  out_dir: "data"             # where the model-ready CSV is written
---

# Overview

This file builds the model-ready longitudinal dataset used by `notebooks/diabetes_onset_lstm.ipynb`.

1. Load static demographics (`sociodemo`), baseline clinical covariates (`common`), office-visit vitals (`followup_crf_office_visit_only`) and time-to-event outcomes (`ttoevent`).
2. Exclude patients with diabetes in the lookback window.
3. Reshape vitals to one row per patient visit and compute time since last visit.
4. Impute missing BMI, systolic and diastolic BP with **multilevel MICE** (patient as cluster). **Outcome variables are excluded from the imputation model** so the outcome cannot leak into the features.
5. Export one CSV for modelling.

All outputs in this document are aggregate. Example trajectories use anonymised labels.

```{r setup, message = FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(mice)
library(naniar)

data_file = function(name) file.path(params$data_dir, name)
```

## Load data
```{r load}
load(data_file("followup_crf_office_visit_only.rdata"))  # dat_crf_office
load(data_file("followup_visit_start_end.rdata"))        # dat_followup_visit_start_end
load(data_file("ttoevent.rdata"))                        # ttoevent
load(data_file("common.rdata"))                          # common
load(data_file("sociodemo.rdata"))                       # sociodemo
```

# Cohort and features

## Exclude prevalent diabetes
```{r exclude}
no_diabetes_lookback_ids = common %>%
  filter(dm_phenotype_lookback == 0) %>%
  pull(pat_id)

dat_crf_office = dat_crf_office %>% filter(pat_id %in% no_diabetes_lookback_ids)
ttoevent = ttoevent %>% filter(pat_id %in% no_diabetes_lookback_ids)
```

## Static features and outcome
```{r static}
labels = ttoevent %>%
  select(pat_id, event_no_pheno, surv_no_pheno) %>%
  rename(diabetes_onset = event_no_pheno)

static_vars = common %>%
  left_join(sociodemo, by = "pat_id") %>%
  select(pat_id, age_years, sex, race_group, asian_ethnic_group, everything()) %>%
  left_join(labels, by = "pat_id")
```

## Time-varying features
```{r dynamic}
tv_vars = dat_crf_office %>%
  select(pat_id, contact_date, bmi_new, bp_systolic_new, bp_diastolic_new) %>%
  arrange(pat_id, contact_date) %>%
  group_by(pat_id) %>%
  mutate(
    time_step = row_number(),
    time_since_last_visit = as.numeric(contact_date - lag(contact_date)),
    time_since_last_visit = replace_na(time_since_last_visit, 0)
  ) %>%
  ungroup()

final_long_data = tv_vars %>%
  left_join(static_vars, by = "pat_id")
```

# Exploratory analysis

## Missingness
```{r missingness}
final_long_data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(missing_percent = round(100 * missing_count / nrow(final_long_data), 2)) %>%
  arrange(desc(missing_percent))
```

## Outcome distribution
```{r outcome}
patient_level = final_long_data %>%
  distinct(pat_id, diabetes_onset, asian_ethnic_group)

patient_level %>%
  count(diabetes_onset) %>%
  mutate(percent = 100 * n / sum(n))

# Groups with fewer than 11 cases are suppressed
patient_level %>%
  group_by(asian_ethnic_group) %>%
  summarise(patients = n(), cases = sum(diabetes_onset, na.rm = TRUE)) %>%
  filter(cases >= 11) %>%
  mutate(percent = 100 * cases / patients) %>%
  ggplot(aes(reorder(asian_ethnic_group, -percent), percent)) +
  geom_col(fill = "#606e4c") +
  geom_text(aes(label = sprintf("%.1f%%", percent)), vjust = -0.5, size = 3.5) +
  labs(title = "Diabetes onset by Asian ethnic group", x = NULL, y = "Converted (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Visit patterns
```{r visits}
visit_summary = final_long_data %>%
  group_by(pat_id) %>%
  summarise(
    n_visits = n(),
    mean_gap = mean(time_since_last_visit[time_step > 1]),
    asian_ethnic_group = first(asian_ethnic_group)
  )

ggplot(visit_summary, aes(n_visits)) +
  geom_histogram(binwidth = 2, fill = "#5c4033", color = "white") +
  labs(title = "Office visits per patient", x = "Visits", y = "Patients") +
  theme_minimal()

ggplot(visit_summary, aes(mean_gap)) +
  geom_histogram(binwidth = 30, fill = "#5c4033", color = "white") +
  facet_wrap(~ asian_ethnic_group, scales = "free_y") +
  labs(title = "Average days between visits by Asian ethnic group", x = "Days", y = "Patients") +
  theme_minimal()
```

# Multilevel imputation (MICE)

BMI, systolic BP and diastolic BP are imputed with `2l.norm`, which models visits nested within patients. Time-invariant covariates use single-level methods. `diabetes_onset` and `surv_no_pheno` are **removed as predictors** so imputed vitals do not encode the outcome.

```{r mice-setup}
imputation_data = final_long_data %>%
  select(
    pat_id, contact_date, time_since_last_visit, surv_no_pheno, diabetes_onset,
    bp_systolic_new, bp_diastolic_new, bmi_new, age_years, sex, asian_ethnic_group,
    privins_lookback, hld_phenotype_lookback, hypertens_lookback, ischhd_lookback,
    cerebrvasd_lookback, heartfail_lookback, perivd_lookback, renald_lookback,
    alcohol_lookback, tobacco_ever_lookback
  ) %>%
  mutate(
    contact_date = as.numeric(contact_date),   # mice cannot handle Date columns
    privins_lookback = as.factor(privins_lookback),
    pat_id_int = as.integer(as.factor(pat_id))
  )

vis_miss(imputation_data, warn_large_data = FALSE)

init = mice(imputation_data, maxit = 0)

method = init$method
method[c("bmi_new", "bp_systolic_new", "bp_diastolic_new")] = "2l.norm"
method["age_years"] = "norm"
method["asian_ethnic_group"] = "polyreg"
method[c("privins_lookback", "sex")] = "logreg"

pred = init$predictorMatrix
pred[, "pat_id_int"] = -2                                   # cluster identifier
pred["pat_id_int", ] = 0
pred[, "pat_id"] = 0
pred["pat_id", ] = 0
pred[, c("diabetes_onset", "surv_no_pheno")] = 0           # no outcome leakage

post = init$post
post["bmi_new"] = "imp[[j]][imp[[j]] < 12] <- 12"          # floor implausibly low BMI
```

```{r mice-run}
mice_result = mice(
  imputation_data,
  m = 5, maxit = 20, seed = 123,
  method = method, predictorMatrix = pred, post = post,
  printFlag = FALSE
)

plot(mice_result)                                                    # convergence
densityplot(mice_result, ~ bmi_new + bp_systolic_new + bp_diastolic_new)
```

## Imputation checks

Observed vs imputed trajectories for five randomly chosen patients with more than three imputed values. Patients are shown as "Patient 1" to "Patient 5"; no identifiers are printed.

```{r mice-check, message = FALSE, warning = FALSE}
vars_to_check = c("bmi_new", "bp_systolic_new", "bp_diastolic_new")
imputed = complete(mice_result, 1)

was_missing = imputation_data %>%
  select(pat_id, contact_date, all_of(vars_to_check)) %>%
  mutate(across(all_of(vars_to_check), is.na))

set.seed(8332)
sample_ids = was_missing %>%
  group_by(pat_id) %>%
  summarise(n_imputed = sum(across(all_of(vars_to_check)))) %>%
  filter(n_imputed > 3) %>%
  slice_sample(n = 5) %>%
  mutate(label = paste("Patient", row_number()))

plot_data = imputed %>%
  inner_join(sample_ids, by = "pat_id") %>%
  select(label, pat_id, contact_date, all_of(vars_to_check)) %>%
  pivot_longer(all_of(vars_to_check), names_to = "variable", values_to = "value") %>%
  left_join(
    was_missing %>% pivot_longer(all_of(vars_to_check), names_to = "variable", values_to = "imputed"),
    by = c("pat_id", "contact_date", "variable")
  ) %>%
  mutate(contact_date = as.Date(contact_date, origin = "1899-12-30"))

ggplot(plot_data, aes(contact_date, value)) +
  geom_line(aes(group = label), color = "#1c1c1c", alpha = 0.4) +
  geom_point(aes(color = imputed, shape = imputed), size = 2) +
  scale_color_manual(values = c(`FALSE` = "#1c1c1c", `TRUE` = "#c35237"),
                     labels = c("Observed", "Imputed"), name = NULL) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17),
                     labels = c("Observed", "Imputed"), name = NULL) +
  facet_grid(variable ~ label, scales = "free") +
  labs(title = "Observed and imputed vitals", x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

# Plausibility: imputed values should fall inside the observed range
sapply(vars_to_check, function(v) {
  obs = imputation_data[[v]]
  imp = imputed[[v]][is.na(obs)]
  c(observed_min = min(obs, na.rm = TRUE), observed_max = max(obs, na.rm = TRUE),
    imputed_min = min(imp), imputed_max = max(imp))
})
```

# Export

The first completed dataset is exported. The visit-level diagnosis label is built in the modelling notebook from `surv_no_pheno`, where post-diagnosis visits are also dropped.

```{r export}
final_imputed_data = imputed %>%
  mutate(contact_date = as.Date(contact_date, origin = "1899-12-30"))

write.csv(final_imputed_data, file.path(params$out_dir, "final_imputed_data.csv"))
```
